# Statistical Learning

## Conceptual

### Question 1

> For each of parts (a) through (d), indicate whether we would generally expect
> the performance of a flexible statistical learning method to be better or
> worse than an inflexible method. Justify your answer.
>
> a. The sample size n is extremely large, and the number of predictors p is
>    small.

Flexible > inflexible
Lots of data available, allowing for a flexible model to capture the underlying patterns with a few predictors.

> b. The number of predictors p is extremely large, and the number of
>    observations n is small.

Inflexible > Flexible
Lots of predictor assigned to little info. May lead to overfitting with a flexible method.
Inflexible method will have higher bias, but lower variance and the predictors will "latch" less spuriously to the data.

> c. The relationship between the predictors and response is highly
>    non-linear.

Flexible > non-nlexible
Flexible methods are more apt to estimate f(x) than inflexible methods, as they have more df to accomodate non-linearity.
e.g., OLS regression vs. adding polynomials or splines

> d. The variance of the error terms, i.e. $\sigma^2 = Var(\epsilon)$, is
>    extremely high.

Inflexible > flexible
Inflexible method less affected by high variability of the error term. Less overfitting through lower variance, higher bias.


### Question 2

> Explain whether each scenario is a classification or regression problem, and
> indicate whether we are most interested in inference or prediction. Finally,
> provide n and p.
>
> a. We collect a set of data on the top 500 firms in the US. For each firm
>    we record profit, number of employees, industry and the CEO salary. We are
>    interested in understanding which factors affect CEO salary.

**Inference** - n = 500; p = 3

> b. We are considering launching a new product and wish to know whether
>    it will be a success or a failure. We collect data on 20 similar products
>    that were previously launched. For each product we have recorded whether it
>    was a success or failure, price charged for the product, marketing budget,
>    competition price, and ten other variables.

**Classification** - n = 20; p = 13

> c. We are interested in predicting the % change in the USD/Euro exchange
>    rate in relation to the weekly changes in the world stock markets. Hence we
>    collect weekly data for all of 2012. For each week we record the % change
>    in the USD/Euro, the % change in the US market, the % change in the British
>    market, and the % change in the German market.

**Prediction**; n = 52; p = 3

### Question 3

> We now revisit the bias-variance decomposition.
>
> a. Provide a sketch of typical (squared) bias, variance, training error,
>    test error, and Bayes (or irreducible) error curves, on a single plot, as
>    we go from less flexible statistical learning methods towards more flexible
>    approaches. The x-axis should represent the amount of flexibility in the
>    method, and the y-axis should represent the values for each curve. There
>    should be five curves. Make sure to label each one.

```{r q3-bias-variance-plot, fig.width=7, fig.height=4.5, echo=TRUE}
library(ggplot2)
library(tidyr)

flex <- seq(0, 1, length.out = 200)

sq_bias   <- 3.8 * (1 - flex)^2.2 + 0.1
variance  <- 0.1 + 3.8 * flex^3
bayes_err <- rep(1.0, length(flex))
test_err  <- sq_bias + variance + bayes_err
train_err <- 3.6 * exp(-3.2 * flex) + 0.2

curves <- data.frame(
  flex = flex,
  `Squared Bias` = sq_bias,
  `Variance`     = variance,
  `Bayes Error`  = bayes_err,
  `Test MSE`     = test_err,
  `Training MSE` = train_err,
  check.names = FALSE
)

opt_flex <- flex[which.min(test_err)]

df_long <- pivot_longer(curves, -flex, names_to = "curve", values_to = "value")
df_long$curve <- factor(df_long$curve, levels = c("Test MSE", "Training MSE",
  "Squared Bias", "Variance", "Bayes Error"))

p <- ggplot(df_long, aes(x = flex, y = value, color = curve, linetype = curve)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = opt_flex, linetype = "dotted", color = "grey40",
    linewidth = 0.8) +
  annotate("text", x = opt_flex + 0.03, y = 5.2, label = "Optimal Flexibility", 
           hjust = 0, size = 3.5, fontface = "italic", color = "grey30") +
  scale_color_manual(
    values = c(
      "Test MSE"     = "#D95F02",
      "Training MSE" = "#7570B3",
      "Squared Bias" = "#1B9E77",
      "Variance"     = "#E7298A",
      "Bayes Error"  = "#333333"
    )
  ) +
  scale_linetype_manual(
    values = c(
      "Test MSE"     = "solid",
      "Training MSE" = "solid",
      "Squared Bias" = "solid",
      "Variance"     = "solid",
      "Bayes Error"  = "dashed"
    )
  ) +
  scale_x_continuous(breaks = c(0, 0.5, 1), labels = c("Low", "Moderate", "High")) +
  scale_y_continuous(breaks = NULL) +
  labs(
    x = "Flexibility",
    y = "Mean Squared Error",
    color = NULL,
    linetype = NULL,
    title = "The Bias-Variance Tradeoff"
  )

if (requireNamespace("ggprism", quietly = TRUE)) {
  p + ggprism::theme_prism(base_size = 13) +
    theme(legend.position = "right", axis.ticks.y = element_blank())
} else {
  p + theme_minimal(base_size = 13) +
    theme(
      axis.line = element_line(color = "black", linewidth = 0.8),
      axis.ticks.x = element_line(color = "black", linewidth = 0.8),
      axis.ticks.y = element_blank(),
      panel.grid = element_blank(),
      legend.position = "right"
    )
}
```


> b. Explain why each of the five curves has the shape displayed in
>    part (a).

**Squared bias** - Inflexible model impose rigid assumptions (e.g., linearity). As flexibility increases, the model as the freedom to bend and approximate complex shapes (f), so bias decreases monotonically.

**Variance** - Increases with flexiblity because the model becomes hypersensitive to the specific training observations and their noise. Fitting a different training set would produce a widly different curve.

**Training error** - Always decreases monotonically because more flexible models have the df to get closer to the training observations (and can interpolate them perfectly).

**Test error** -  Follows a U-shape because $test MSE = bias^2 + Var(f) + \epsilon^2$. Initially, bias drops faster than variance rises (error goes down). But past the optimal point, Var(f) explodes faster than bias decreases, driving test error back up.

**Irreducible error** - Stays flat because it stems from unmeasured variables, sensoir noise and random fluctuatioi that do not depend on the model or its flexibility.


### Question 4

> You will now think of some real-life applications for statistical learning.
>
> a. Describe three real-life applications in which classification might
>    be useful. Describe the response, as well as the predictors. Is the goal of
>    each application inference or prediction? Explain your answer.

**1. Bank churn:** 
Response = Churned within 6 months (Yes/No). 
Predictors = account balance, transaction volume, tenure.
Goal = _Prediction_ (flagging accounts to offer promotions)

**2. Chronic pain:** 
Response = Pain resolved (Yes/No).
Predictors = demographics, pain sensitization score.
Goal = _Inference_ (understanding drivers of recovery).

**Employee turnover:**
Response = Resigned within 1 year (Yes/No).
Predictors = salary, satisfaction score, commute time.
Goal = _Inference_ (designing retention policies)

> b. Describe three real-life applications in which regression might be
>    useful. Describe the response, as well as the predictors. Is the goal of
>    each application inference or prediction? Explain your answer.

**Pain trajectories:**
Response = Continuous pain rating (e.g., 0-10 numerical rating scale).
Predictors = session count, exercise intensity, baseline pain.
Goal = _Inference/Prediction_

**Blood pressure:**
Response = Change in systolic BP (mmHg).
Predictors = treatment dosage, age, baseline BP.
Goal = _Inference_ (estimating treatment efficacy).

**Test MSE:**
Response = Numerica test MSE.
Predictors = degrees of freedom/model frexibility.
Goal: _Prediction_

> c. Describe three real-life applications in which cluster analysis might be
>    useful.

**Cancer cells:**
If we have tumor samples and no labels, we can use clustering to discover novel cancer subtypes based on gene expression profiles.

**Stars:**
Grouping astronomical bodies based on luminosity, temperature and spectral pattern witout predifined categories.

**Car insurance**:
Customer segmentation -- grouping policy-holders into driving profiles (e.g., weekend drivers, high-speed commuters, rural drivers) based on telemetric data to tailor marktering or product tiers.


### Question 5

> What are the advantages and disadvantages of a very flexible (versus a less
> flexible) approach for regression or classification? Under what circumstances
> might a more flexible approach be preferred to a less flexible approach? When
> might a less flexible approach be preferred?

**Flexible model:**
Will be prefered when the goal is _prediction_, i.e., when we're more interested in accurately prediction Y than understanding f(x). Because of the low bias, it should be able to match the true, underlying shape more easily. However, because it has high variance, it is at higher risk of overfitting, especially if the number of observation is small.

**Inflexible model:**
Will be prefered when the goal is _inference_, when we're willing to sacrifice some accuracy in Y (high bias) to be able to understand f(x). Because of the low variance, is less at risk of overfitting, but will not capture the true, underlying shape as easily as a flexible model would. However, f(x) becomes easier to interpret.


### Question 6

> Describe the differences between a parametric and a non-parametric statistical
> learning approach. What are the advantages of a parametric approach to
> regression or classification (as opposed to a non-parametric approach)? What
> are its disadvantages?

**Parametric:**
- Reduces the problem to estimating a fixed, finite set of parameters. Once parameters are known, training data can be discarded.
- _Advantages:_ Much easier to estimate, requires far fewer observationis, less risky for overfitting, highly interpretatable.
- _Disadvantages:_ High bias if the assumed function form (e.g., straight line) doesn't match the true form.

**Non-parametric:**
- Makes no explicit assumption about the functional form of f(x) and does not fit a fixed set of parameters. Instead, lets the data directly define the shape.
- _Advantages:_ Can fit almost any arbitrary, complex shape (very low bias)
- _Disadvantages:_ Requires a large number of observations to avoid overfitting, and is computationally heavy and harder to interpret

### Question 7

> The table below provides a training data set containing six observations,
> three predictors, and one qualitative response variable. 
> 
> | Obs. | $X_1$ | $X_2$ | $X_3$ | $Y$   |
> |------|-------|-------|-------|-------|
> | 1    | 0     | 3     | 0     | Red   |
> | 2    | 2     | 0     | 0     | Red   |
> | 3    | 0     | 1     | 3     | Red   |
> | 4    | 0     | 1     | 2     | Green |
> | 5    | -1    | 0     | 1     | Green |
> | 6    | 1     | 1     | 1     | Red   |
> 
> Suppose we wish to use this data set to make a prediction for $Y$ when 
> $X_1 = X_2 = X_3 = 0$ using $K$-nearest neighbors.
>
> a. Compute the Euclidean distance between each observation and the test
>    point, $X_1 = X_2 = X_3 = 0$.

```{r}
obs <- rep(0, 3)

distance_1 <- sqrt(sum((c(0, 3, 0) - obs)^2))
distance_2 <- sqrt(sum((c(2, 0, 0) - obs)^2))
distance_3 <- sqrt(sum((c(0, 1, 3) - obs)^2))
distance_4 <- sqrt(sum((c(0, 1, 2) - obs)^2))
distance_5 <- sqrt(sum((c(-1, 0, 1) - obs)^2))
distance_6 <- sqrt(sum((c(1, 1, 1) - obs)^2))

kkn_vector <- c(N1 = distance_1, N2= distance_2, N3 = distance_3, N4 = distance_4, N5 = distance_5, N6 = distance_6)
sort(kkn_vector)

```

>
> b. What is our prediction with $K = 1$? Why?

```{r}
kkn1 <- sort(kkn_vector)[1]
```

Observation 5 is the closest neighbor, which is **Green**.

>
> c. What is our prediction with $K = 3$? Why?

```{r}
kkn3 <- sort(kkn_vector)[1:3]
```

Observation 5 (_green_), 6 (_red_) and 2 (_red_) are the closest neighbors. Majority vote wins, prediction is **red**.

>
> d. If the Bayes decision boundary in this problem is highly non-linear, then
>    would we expect the best value for $K$ to be large or small? Why?

If the true boundary is highly non-linear, we need a flexible model to bend and capture those curves. Therefore, K should be small.

```{r}
dat <- data.frame(
  "x1" = c(0, 2, 0, 0, -1, 1),
  "x2" = c(3, 0, 1, 1, 0, 1),
  "x3" = c(0, 0, 3, 2, 1, 1),
  "y" = c("Red", "Red", "Red", "Green", "Green", "Red")
)
```

## Applied

### Question 8

> This exercise relates to the `College` data set, which can be found in
> the file `College.csv`. It contains a number of variables for 777 different
> universities and colleges in the US. The variables are
>
> * `Private` : Public/private indicator
> * `Apps` : Number of applications received
> * `Accept` : Number of applicants accepted
> * `Enroll` : Number of new students enrolled
> * `Top10perc` : New students from top 10% of high school class
> * `Top25perc` : New students from top 25% of high school class
> * `F.Undergrad` : Number of full-time undergraduates
> * `P.Undergrad` : Number of part-time undergraduates
> * `Outstate` : Out-of-state tuition
> * `Room.Board` : Room and board costs
> * `Books` : Estimated book costs
> * `Personal` : Estimated personal spending
> * `PhD` : Percent of faculty with Ph.D.'s
> * `Terminal` : Percent of faculty with terminal degree
> * `S.F.Ratio` : Student/faculty ratio
> * `perc.alumni` : Percent of alumni who donate
> * `Expend` : Instructional expenditure per student
> * `Grad.Rate` : Graduation rate
>
> Before reading the data into `R`, it can be viewed in Excel or a text
> editor.
>
> a. Use the `read.csv()` function to read the data into `R`. Call the loaded
>    data `college`. Make sure that you have the directory set to the correct
>    location for the data.
>
> b. Look at the data using the `View()` function. You should notice that the
>    first column is just the name of each university. We don't really want `R`
>    to treat this as data. However, it may be handy to have these names for 
>    later. Try the following commands:
>    
>    ```r
>    rownames(college) <- college[, 1]
>    View(college)
>    ```
>    
>    You should see that there is now a `row.names` column with the name of 
>    each university recorded. This means that R has given each row a name 
>    corresponding to the appropriate university. `R` will not try to perform 
>    calculations on the row names. However, we still need to eliminate the 
>    first column in the data where the names are stored. Try
>    
>    ```r
>    college <- college [, -1]
>    View(college)
>    ```
>    
>    Now you should see that the first data column is `Private`. Note that 
>    another column labeled `row.names` now appears before the `Private` column.
>    However, this is not a data column but rather the name that R is giving to
>    each row.
>
> c.
>     i.   Use the `summary()` function to produce a numerical summary of the
>          variables in the data set.
>     ii.  Use the `pairs()` function to produce a scatterplot matrix of the
>          first ten columns or variables of the data. Recall that you can 
>          reference the first ten columns of a matrix A using `A[,1:10]`.
>     iii. Use the `plot()` function to produce side-by-side boxplots of 
>          `Outstate` versus `Private`.
>     iv. Create a new qualitative variable, called `Elite`, by _binning_ the
>         `Top10perc` variable. We are going to divide universities into two
>         groups based on whether or not the proportion of students coming from
>         the top 10% of their high school classes exceeds 50%.
>         
>         ```r
>         > Elite <- rep("No", nrow(college))
>         > Elite[college$Top10perc > 50] <- "Yes"
>         > Elite <- as.factor(Elite)
>         > college <- data.frame(college, Elite)
>         ```
>         
>         Use the `summary()` function to see how many elite universities there
>         are. Now use the `plot()` function to produce side-by-side boxplots of
>         `Outstate` versus `Elite`.
>     v.   Use the `hist()` function to produce some histograms with differing
>          numbers of bins for a few of the quantitative variables. You may find
>          the command `par(mfrow=c(2,2))` useful: it will divide the print 
>          window into four regions so that four plots can be made 
>          simultaneously. Modifying the arguments to this function will divide
>          the screen in other ways.
>     vi.  Continue exploring the data, and provide a brief summary of what you
>          discover.

### Question 9

> This exercise involves the Auto data set studied in the lab. Make sure
> that the missing values have been removed from the data.
>
> a. Which of the predictors are quantitative, and which are qualitative?
>
> b. What is the range of each quantitative predictor? You can answer this using
>    the `range()` function.
>
> c. What is the mean and standard deviation of each quantitative predictor?
>
> d. Now remove the 10th through 85th observations. What is the range, mean, and
>    standard deviation of each predictor in the subset of the data that 
>    remains?
>
> e. Using the full data set, investigate the predictors graphically, using
>    scatterplots or other tools of your choice. Create some plots highlighting
>    the relationships among the predictors. Comment on your findings.
>
> f. Suppose that we wish to predict gas mileage (`mpg`) on the basis of the
>    other variables. Do your plots suggest that any of the other variables
>    might be useful in predicting `mpg`? Justify your answer.

### Question 10

> This exercise involves the `Boston` housing data set.
>
> a. To begin, load in the `Boston` data set. The `Boston` data set is part of 
>    the `ISLR2` library in R.
>    
>    ```r
>    > library(ISLR2)
>    ```
>    
>    Now the data set is contained in the object `Boston`.
>    
>    ```r
>    > Boston
>    ```
>    
>    Read about the data set:
>    
>    ```r
>    > ?Boston
>    ```
>    œ
>    How many rows are in this data set? How many columns? What do the rows and
>    columns represent?
>
> b. Make some pairwise scatterplots of the predictors (columns) in this data
>    set. Describe your findings.
>
> c. Are any of the predictors associated with per capita crime rate? If so,
>    explain the relationship.
>
> d. Do any of the census tracts of Boston appear to have particularly high 
>    crime rates? Tax rates? Pupil-teacher ratios? Comment on the range of each
>    predictor.
>
> e. How many of the census tracts in this data set bound the Charles river?
>
> f. What is the median pupil-teacher ratio among the towns in this data set?
>
> g. Which census tract of Boston has lowest median value of owner-occupied
>    homes? What are the values of the other predictors for that census tract,
>    and how do those values compare to the overall ranges for those predictors?
>    Comment on your findings.
>
> h. In this data set, how many of the census tract average more than seven
>    rooms per dwelling? More than eight rooms per dwelling? Comment on the
>    census tracts that average more than eight rooms per dwelling.
